|

### 2022-04-15 article/article.Rmd
add large text revision

---
title:  |
        | Legislature Size and Welfare: Evidence from Brazil
author: 
date: \today
abstract: "How does legislature size impact public service provision? Despite the importance of institutional design for democratic governance, the effect of particular legislative features on citizen welfare remains little understood. In this paper, we use a formal model to show that increasing legislature size improves public goods provision. We argue that changes in bargaining costs depend on whether additional legislators share the Executive's party affiliation: More opposition members reduce the Executive's public goods provision, while more government-aligned members increase it. We test this theory by exploiting sharp discontinuities in city council size in Brazil. We show that an additional city councilor has a 91% chance of belonging to the mayoral coalition, and this significantly improves primary school enrollment and infant mortality measures. To explore possible mechanisms, we surveyed former councilors and analyzed 346,553 bills proposed between 2005 and 2008. This paper has implications for the design of representative institutions."
abstractspacing: double
keywords: Legislature Size; Legislative Bargaining; Local Public Goods Provision; Regression Discontinuity Design; Brazil
jelcodes: H41; H80; I31; I38
fontsize: 12pt
margin: 2cm
urlcolor: darkblue
linkcolor: Mahogany
citecolor: Mahogany
spacing: double
papersize: a4paper
bibliography: references.bib
biblio-style: apalike
output:
  pdf_document:
    citation_package: natbib
    fig_caption: yes
    number_sections: yes
    keep_tex: no
    toc: no
    toc_depth: 3
    template: article-template.latex
    md_extensions: +raw_attribute
---

@@

Legislative institutions are crucial for citizen welfare and service provision in democratic countries. Historically, the division of power and the birth of a representative government placed legislatures at the center of the political stage. In modern democracies, most countries have assemblies at all levels of government. These institutions regulate the quantity and quality of legislation, check and veto other political powers, and influence policymaking. Moreover, legislatures also determine taxation, government spending and oversee public programs \citep{auriol2012optimal, north1989constitutions, poulsen2018political, stasavage2011states, Weingastetal1981, weingast1988industrial}.

Regardless of eventual differences in electoral rules, social composition, and internal procedures, all legislatures share one feature: they are collective bodies. As such, the number of representatives has important consequences for the whole polity. Indeed, legislature size is known to have a significant effect on collective action strategies, the bargaining power between state branches, government spending, and representation \citep{crain1979cost, freire2021legislature, PrimoSnyder2008, rogers2002free, Weingastetal1981}. However, how legislature size impacts social welfare remains less understood. On the one hand, enlarged legislatures may increase government spending and lawmaking, which can lead to more public goods \citep{Weingastetal1981}. On the other hand, adding a legislator may increase the number of veto players and exacerbate collective action problems, thus hindering service provision \citep{gottlieb2019countervailing, tsebelis2002veto}. Therefore, what is the effect of larger legislatures on public service provision?

In this paper, we propose a local legislature size and welfare theory that incorporates two crucial features: Executive leadership and party politics. First, in mayor-council municipalities, mayors tend to be central in local public service decisions. Mayors have the prerogative to boost public investments, like presidents in federal governments. Second, partisanship is essential at the local level, even in settings where the ideological attachment is weak, such as in most young democracies. Party attachment structures political views \citep{samuels2014power}, and local-level elites usually belong to different factions and parties. They fight for power in the municipality by controlling various parties and forming distinct pre-electoral coalitions.\footnote{Elite political competition in the local level has also been linked to violence and conflict in Brazil and elsewhere \citep{hoelscher2015politics, mignozzetti2018predatory}.} In many cities, family ties may be substitutes for the ideological component, but influential local leaders usually sort themselves into different political groups \citep{ferraz2020political, nunes1977coronelismo}.

This paper incorporates partisanship and mayoral first-move advantage to a \citet{baron1989bargaining} bargaining model. Mayors make the initial policy proposal that consists of an investment in public goods provision, a decision to extract rents for personal gains, and a vector of side-transfers for councilors. If they fail to secure majoritarian support from councilors, a reversal stage occurs, where the decision leaves the mayor's office, and councilors propose the public investments. In this mayor-council bargaining process, costs come from the side-transfers required to secure a majority among councilors. We show that in a model without partisan concerns, bargaining costs always increase with larger city councils, which in turn lower public goods provision and rents. However, partisanship decreases bargaining costs when the chances of electing councilors aligned with the mayor are sufficiently high. Lower bargaining costs release funds for mayors to invest in public goods provision. In this case, a larger city council may have more partisan councilors aligned with the mayor, reducing the mayor's bargaining costs and increasing the equilibrium public goods provision.

We test these theoretical predictions by exploiting an exogenous variation in Brazilian city council sizes. In 2004, the judiciary reinterpreted the Brazilian Constitution and unexpectedly changed the number of councilors in several municipalities. Until then, towns could freely determine their legislature size, but in March 2004, the Supreme Electoral Court created population thresholds and added one councilor to each set of 47,619 inhabitants. Around these cutoffs, the decision represented an exogenous change in city council size, allowing us to investigate the effects of increasing the number of local legislators in Brazil.

We find that larger city councils changed the composition of mayoral coalitions and the number of appointed bureaucrats in the municipality. The extra city councilor had a 91 percent chance of belonging to the mayor's pre-electoral coalition. This enhanced the chances of electing a councilor connected with the mayor, which then decreased the bargaining costs for the mayor and boosted investments in public services. Moreover, the number of appointed bureaucrats increased by 104 with the additional councilor. In Brazil, appointing bureaucrats leads to higher quality workers, better service provision, and more accountability \citep{toral2019benefits}.

Higher mayor support and more appointed bureaucrats correlate with city councilors' service provision expectations and legislation approved in Brazilian municipalities. We surveyed 174 former councilors who served during the 2005-2008 term, and they corroborate that mayors use bureaucratic appointments and political favors to secure support. As the formal model shows, bargaining with councilors requires offering transfers to the representatives. By investigating what councilors prefer to provide to voters, we find that councilors mainly demand job appointments, as these increase their reelection chances and improve municipal welfare. These findings align with a novel dataset we built with 346,553 laws passed in 63 municipalities within ten thousand inhabitants away from the cutoffs. The legislation in municipalities on the left and on the right-hand side of the cutoffs mentioned local public goods. However, municipalities with larger city councils had around 15 percent more public goods provision proposals, which is in line with our expectations.

Finally, we show that the increase in legislature size significantly impacts municipal education and health care. Increasing the city council size by one legislator lowers infant mortality by 2.01 children per one thousand infants born and reduces postneonatal mortality by 0.90 children per one thousand infants that survived their first 28 days. A larger council size also increased enrollment by 2.58 children in elementary education classrooms without significantly diminishing elementary school quality. These services are crucial for welfare and result from eased bargaining between the mayor and the city council.

This paper contributes to four strands of the literature. First, this paper dialogues with a large literature on multilateral bargaining. Repeated models of multilateral bargaining started with \citet{baron1989bargaining}, who extended the bilateral bargaining game proposed by \citet{rubinstein1982perfect} to more than two players. Several papers followed \citet{baron1989bargaining}, manipulating endogenous recognition power, the size of the required majority, and partisanship in the models \citep{agranov2020persistence, baranski2020determinants, calvert2005legislative, choate2019partisan}.\footnote{See \citet{eraslan2019legislative} for a comprehensive review of multilateral bargaining and \citet{ray2015coalition} for a review on coalition formation.} In this paper, we extend the \citet{baron1989bargaining} model under closed rules and add the mayor's first move and political partisanship. Partisanship here is different from \citet{choate2019partisan} since we consider partisanship as a zero-sum effect that is positive for politicians aligned with the mayor and negative otherwise. This paper investigates the often overlooked effect of legislature size on the bargaining equilibrium strategies.

Second, this paper contributes to the literature on the effects of larger legislatures in service provision. Following the contribution by \citet{Weingastetal1981}, the "law of $1/n$" posits that larger legislatures increase the legislative pressure for providing targeted public goods to local supporters at the expense of the entire polity \citep{freire2021legislature}. The corollary of the "law of $1/n$" is that increasing legislature size promotes excessive public spending.\footnote{\citet{PrimoSnyder2008} show that legislature size may result in positive or negative expenditures, depending on factors such as the degree that local public goods can exclude non-locals or deadweight costs associated with taxation.} In the US metaphor, this would explain significant infrastructure spending in smaller districts. However, the "law of $1/n$" literature rarely mentions the relationship between public goods provision and welfare. This paper is one of the first to study how the number of local legislators affects welfare. Our model provides a theoretical expectation based on the chances of electing a politician aligned with the mayor. The empirical estimates confirm our theoretical predictions, with the additional fact that legislature size has a null effect on expenditures in Brazil.

Third, our article also contributes to a growing literature on municipal partisan politics. Most papers in this area estimate the effects of the mayor's party on political outcomes. For instance, \citet{de2016mayoral} and \citet{gerber2011mayors} find differences in governance when comparing Democratic and Republican mayors. Specifically about Brazil, \citet{gouvea2021partisanship} shows that social policy expenditure increase with left-wing mayors. \citet{frey2020larger}, using discontinuities in city council size after the 2009 constitutional amendment that reverted the thresholds we study here, shows that changes in council size lower the support for candidates backed by the municipal mayors. Our paper reinforces the view that partisanship matters locally, even in countries with weak ideological linkage.

Finally, this paper contributes to a growing literature on the causes and effects of legislature sizes. Most papers on legislature size study the trade-off between representation and efficiency \citep{frederick2008people, jacobs2015explaining, muzzio1989size, Stigler1976}. However, some recent articles explore the effects of legislature size on lawmaking and partisanship. \citet{kirkland2014chamber} shows that larger legislatures make it harder for politicians to know each other and collaborate, making them rely more heavily on party cues and committee structures. In considering the causes of changes in legislature sizes, \citet{jacobs2015explaining} show that population shifts and economic crises cause enlargement and shrinkage, respectively. \citet{marland2019fewer} questions the economic crises mechanism, showing that political elites shrink legislatures mostly to signal to the public the priority of their austerity agenda. In this article, we show that legislature size affects welfare by changing the partisan composition of legislatures.

@@

The design of institutions is crucial for long term welfare \citep{acemoglu2012nations, buenodemesquita2005logic, north1989constitutions} and democratic consolidation \citep{gandhi2019institutional, lijphart2018institutional}. One particular problem that each polity faces is determination of the size of its legislative bodies. The literature is extensive, and modern attempts to provide a solution to the legislature size problem started as early as in the eighteenth century, with the Federalists versus Anti-Federalists debates \citep{muzzio1989size, Stigler1976}.

The legislature size literature highlights that the number of legislators causes tension between representation and efficiency \citep{frederick2008people, jacobs2015explaining, muzzio1989size, rush2013conclusion, Stigler1976, taagepera1972size}. Larger legislatures improve representation as the election thresholds decrease and the ratio of population-per-politician decreases. This facilitates the electoral success of ethnic and political minorities \citep{allenrepresentation}. Conversely, the improvements in representation are countered by the decline in efficiency in larger legislatures. The enlarged size makes it harder for members to coordinate and collaborate without partisanship and legislative structures \citep{kirkland2014chamber, rush2013conclusion}. It also increases the potential number of veto players, decreasing the policymaking efficiency \citep{tsebelis2002veto}. Balancing these two incentives is crucial for the optimal legislative composition \citep{auriol2012optimal, taagepera1972size}.

However, the legislature size scholarship has overlooked two questions: first, does legislature size matter in contexts where the legislative bodies are not the pivotal decision-maker? And second, what is the ultimate effect of legislature size on welfare? On the one hand, legislature size's effects on welfare represent a significant gap in the political economy literature. It is likely generated by the predominance of the "law of 1/n" scholarship, which emphasized larger legislatures' overspending consequence. Since the seminal paper by \citet{Weingastetal1981}, 30 papers have systematically tested the effects of legislature size on government spending \citep{freire2021legislature}. Of these papers, only \citet{lewis2019legislature} investigates its effects on welfare, showing that larger legislatures have a negative welfare effect.\footnote{\citet{koppel2004public} also shows, using a repeated game where the proposal prerogative rotates in a legislature, that larger legislatures increase the chances of non-cooperative outcomes, lowering public goods provision.} \citet{frey2020larger} analyses the Brazilian case after the 2009 legislature size change and finds a null effect in service provision from 2012 to 2016. These results suggest that the legislature size and welfare nexus needs a dedicated theoretical and empirical assessment.

Despite the division of power, the Executive dominates the debates on public service provision in most mayor-council (or presidential) systems. For instance, while the New York Times cited the New York City Mayor Bill De Blasio 2,084 times in 2021, it mentioned the 49 seating city councilors 440 times in the same period. To avoid the COVID crisis when the Executive was in the spotlight, we expanded the search for the past five years. In this new search, Bill De Blasio was mentioned 8,978 while the 49 seating city councilors were mentioned 2,046 times.\footnote{These figures were computed using the Dow Jones Factiva Database in November 2021.} The mayor was cited, on average, over four times more than all the city councilors combined. This Executive dominance has implications for policymaking, as coalitions and affinity with the Executive are going to be crucial for service provision \citep{buenodemesquita2005logic}.

@@

The city council votes the mayor's proposal, and if councilors accept it, the proposal is implemented. Otherwise, a reversal policy takes place. Reversal policy in this context means that the mayor's proposal failed to secure the majority of council votes. The decision process then leaves the mayor's office and it is transferred to the council. The types of reversal policies are crucial for our argument, as it changes the relative strength of the council when they bargain with the mayor. We investigate two types of reversal policies: a baseline reversal with no parties and a hybrid mechanism with partisan and non-partisan concerns. The differences between these two mechanisms are that the non-partisan councilors only care about rents, while partisan councilors care about both rents and policy.

The mayor must convince at least half of the city councilors to support her policy proposals. Similar to \citet{buenodemesquita2005logic}, we assume that the selectorate, that is, the individuals who have a say in the mayor's policy proposals, are the $N$ city councilors. The winning coalition, the minimum number of members that must be convinced to implement the policy, is equal to half of the selectorate ($N/2$). City councilors are motivated by policy concerns $p$ and transfers received from the mayor or the councilor nominated as the proposer \textbf{x}. We leave the definition of these transfers open, as this allows us to analyze how mayors combine transfers with other incentives to pressure city councilors. Exemples of transfers are granting electoral favors, favoring clientelistic brokers, spending in areas that the councilor has political or personal interests, providing portfolio (municipal offices) and public jobs within the municipal bureaucracy, or even giving campaign contributions and bribes. The use of transfers to influence the councilors' choices generate a bargaining cost of $C_G$. Bargaining costs vary according to the reversal mechanism and the size of the legislature.

Finally, after learning the bargaining costs, the mayor provides a level of public goods provision $g$ and pockets rents $r$. Public goods provision helps the mayor to get reelected. Rents are for direct mayor's consumption, and they do not contribute to the mayor's electoral success.\footnote{This model, similar to \citet{cowen1994rent} and \citet{buenodemesquita2005logic}, combines public goods provision and rent extraction. For a dynamic bargaining model that incorporates transfers to districts and public goods provision, see \citet{battaglini2012legislative}.} This makes the mayor's expected utility a sum of the gains from rents and the benefits from reelection. Both the utilities from rents $u(.)$ and the probability of reelection $\pi(.)$ are concave functions, meaning that more rents or public goods increase the utility at a decreasing rate. The probability of reelection is multiplied by the benefit from holding office $B_M > 0$. This benefit captures the tangible and intangible gains that the mayor receives from holding the public office. We assume that the benefits are high enough to rule out an equilibrium with only rents and transfers, which is equivalent to saying that mayors strongly prefer to be reelected. If the mayor's policy proposal is approved, then the expected utility for the mayor is:

@@

The expected utility for the city councilors depends on the reversal policy. In the baseline non-partisan reversal, when the council rejects the mayor's proposal, a reversal stage starts with resources diminishing by a factor of $\delta \in (0,1)$ and the selection of one councilor randomly. The selected councilor becomes the new proposer. If her proposal is accepted, it is implemented. If councilors reject the offer again, the budget shrinks by $\delta$, and another city councilor is nominated to propose. The process repeats until a proposal is finally accepted.

We now consider a situation that combines both partisan and non-partisan legislature concerns. Each councilor has a party affiliation. Party affiliations are mutually exclusive, and if a city councilor is aligned with the mayor's party, then she belongs to the government coalition $G \subset \{1, 2, \cdots, N \}$. Otherwise, the city councilor belongs to the opposition $O = \{1, 2, \cdots, N \} \setminus G$. The mayor's policy generates a political value of $p > 0$ for the councilors, and whether it adds or subtracts from the councilor's utility depends on the party affiliation of the councilor.\footnote{One example of partisan benefits is that, as the mayor shares partisan ties with members of the government coalition, a fraction of the public goods provision can be claimed by councilors as their contribution. We propose an extension in the Full Supplementary Materials (available in [REDACTED FOR PEER REVIEW]) that assumes $p = \phi g$, where $\phi$ is the fraction of the public goods provision that can be credit claimed by the city councilor aligned with the mayor.} If councilors reject the mayor's proposal, no public goods are provided ($p=0$). Moreover, in equilibrium, we assume that city councilors would prefer to receive transfers instead of providing public goods.\footnote{In the Full Supplementary Materials, we consider a version of the model where public goods provision depends on the party affiliation of the proposer so that it affects choices in any potential history of the game. The equilibrium public goods provision level is even higher than in the model presented here, as city councilors prefer to approve the mayor's proposal than run the risk of an opposition member becoming the proposer.}

@@

Proposition \ref{prop1} provides the first empirical hypothesis to test: \textit{if the bargaining costs decrease when the legislature size increases, then public goods provision increases when the legislature size increase}. Now we study each reversal mechanism to determine the condition under which the bargaining costs decrease when legislature size increases.

@@

\begin{proposition}
In the baseline non-partisan reversal, bargaining costs always increase when the size of the legislature increases.
\end{proposition}

>>>>

---
title:  |
        | Legislature Size and Welfare: Evidence from Brazil
author: 
date: \today
abstract: "How does legislature size impact public service provision? Despite the importance of institutional design for democratic governance, the effect of legislative features on citizen welfare remains little understood. In this paper, we use a formal model to show that increasing legislature size improves public goods delivery. We argue that changes in bargaining costs depend on whether additional legislators share the executive's party affiliation: More opposition members reduce the executive's public goods provision, while more government-aligned members increase it. We test this theory by exploiting sharp discontinuities in city council size in Brazil. We show that an additional city councilor has a 91% chance of belonging to the mayoral coalition, and this significantly improves primary school enrollment and infant mortality measures. To explore possible mechanisms, we surveyed former city councilors and analyzed 346,553 bills proposed between 2005 and 2008. This paper has implications for the design of representative institutions."
abstractspacing: double
keywords: Legislature Size; Legislative Bargaining; Local Public Goods Provision; Regression Discontinuity Design; Brazil
jelcodes: H41; H80; I31; I38
fontsize: 12pt
margin: 2cm
urlcolor: darkblue
linkcolor: Mahogany
citecolor: Mahogany
spacing: double
papersize: a4paper
bibliography: references.bib
biblio-style: apalike
output:
  pdf_document:
    citation_package: natbib
    fig_caption: yes
    number_sections: yes
    keep_tex: no
    toc: no
    toc_depth: 3
    template: article-template.latex
    md_extensions: +raw_attribute
---

@@

Legislative institutions are crucial for citizen welfare and public service
provision in democratic countries. Historically, the division of power and the
birth of representative governments placed legislatures at the center of the
political stage. In modern democracies, most countries have assemblies at all
levels of government. These institutions regulate the quantity and quality of
legislation, check and veto other political powers, and influence policymaking.
Moreover, legislatures also determine taxation, government spending and oversee
public programs \citep{auriol2012optimal, north1989constitutions,
poulsen2018political, stasavage2011states, Weingastetal1981}.

Regardless of eventual differences in electoral rules, social composition, and
internal procedures, all legislatures share one common feature: They are
collective bodies. As such, the number of elected representatives has important
consequences for the whole polity. Legislature size is known to have a
significant effect on collective action strategies, the bargaining power
between state branches, government spending, and representation
\citep{crain1979cost, freire2021legislature, gerring2020population,
PrimoSnyder2008, rogers2002free, Weingastetal1981}. However, how legislature
size impacts social welfare remains less understood. On the one hand, enlarged
legislatures may reduce political particularism and increase overall government
spending, which can lead to more public goods \citep{gerring2020population,
GilliganMatsusaka2001, Weingastetal1981}. On the other hand, adding more
representatives to a legislature may exacerbate collective action problems,
thus hindering service provision \citep{gottlieb2019countervailing,
tsebelis2002veto}. Therefore, what is the effect of larger legislatures on
public service delivery?

In this paper, we propose a new theory based on executive leadership and party
politics to explain how city council size impacts local public goods provision.
First, in mayor-council municipalities, mayors are the main responsible for
public service decisions and have the power to boost public investment similar
to presidents and governors. Second, partisanship plays a key role in city
politics, even in countries in which ideological attachment is relatively weak.
Party attachment structures political views \citep{samuels2014power}, and
local-level elites usually belong to different factions and parties. They fight
for power in the municipality by controlling various parties and forming
distinct pre-electoral coalitions.\footnote{Elite political competition in the
local level has also been linked to violence and conflict in Brazil and
elsewhere \citep{hoelscher2015politics, mignozzetti2018predatory}.} In many
cities, family ties may be substitutes for the ideological component, but
influential local leaders usually sort themselves into different political
groups \citep{ferraz2020political, nunes1977coronelismo}.

This paper incorporates partisanship and mayoral first-move advantage to a
\citet{baron1989bargaining} bargaining model. Mayors make the initial policy
proposal, which consists of investments in public goods provision, rents for
personal gains, and a vector of side-transfers for councilors. If mayors fail
to secure majoritarian support from councilors, a reversal stage occurs, in
which the decision leaves the mayor's office and councilors propose the public
investments. In this mayor-council bargaining process, costs come from the
side-transfers required to secure a majority among councilors. We show that in
a model without partisan concerns, bargaining costs always increase with larger
city councils, which in turn lower public goods provision and rents. However,
partisanship decreases bargaining costs when the chances of electing councilors
aligned with the mayor are sufficiently high. Lower bargaining costs release
funds for mayors to invest in public goods provision. In this case, a larger
city council may have more partisan councilors aligned with the mayor, reducing
the mayor's bargaining costs and increasing the equilibrium public goods
provision.

We test these theoretical predictions by exploiting exogenous variation in city
council size in Brazil. In 2004, the judiciary reinterpreted the Brazilian
Constitution and unexpectedly changed the number of councilors in several
municipalities. Until then, cities could freely determine their legislature
size, but in March 2004 the Supreme Electoral Court created population
thresholds and added one councilor to each set of 47,619 inhabitants. Around
these cutoffs, the decision represented an exogenous change in city council
size, allowing us to investigate the effects of increasing the number of local
legislators in Brazil.

We find that larger city councils changed the composition of mayoral coalitions
and the number of appointed bureaucrats in the municipality. The extra city
councilor had a 91 percent chance of belonging to the mayor's pre-electoral
coalition. With a larger coalition, mayors face lower bargaining costs and are
able to boost public investments with fewer constraints. Each additional
legislator also leads to 104 new appointed bureaucrats, and research shows that
political appointees civil servants improve public service provision and
enhance accountability in Brazil \citep{toral2019benefits}.

To provide further evidence for our mechanisms, we surveyed 174 former
councilors who served during the 2005-2008 term, the period we analyse in this
paper. They confirm that mayors indeed use bureaucratic appointments and
political favors to secure legislative support. As our formal model shows,
bargaining with councilors requires offering transfers to the representatives.
By asking councilors what they would like to provide to voters, we find that
councilors mainly demand job appointments, as these increase their reelection
prospects and improve municipal welfare. These findings are in line with
quantitative results from a novel dataset we built with 346,553 laws passed in
63 municipalities within ten thousand inhabitants away from the cutoffs. We see
that legislation in municipalities both on the left- and on the right-hand side
of the cutoffs mention public goods provision. However, municipalities with
larger city councils have 15 percent more proposals about public goods
provision, which is consistent with our argument.

Finally, we show that the increase in legislature size significantly impacts
municipal education and health care. Increasing the city council size by one
legislator lowers infant mortality by 2.01 children per one thousand infants
born and reduces postneonatal mortality by 0.90 children per one thousand
infants that survived their first 28 days. Larger councils also increase
enrollment by 2.58 children in elementary education classrooms without reducing
school quality. These services are crucial for citizen welfare and result from
eased bargaining between the mayor and the city council.

This paper contributes to four strands of the literature. First, this paper
dialogues with a large literature on multilateral bargaining. Repeated models
of multilateral bargaining started with \citet{baron1989bargaining}, who
extended the bilateral bargaining game proposed by
\citet{rubinstein1982perfect} to more than two players. Several papers followed
\citet{baron1989bargaining} and included endogenous recognition power, the size
of the required majority, and partisanship in the model
\citep{agranov2020persistence, baranski2020determinants,
calvert2005legislative, choate2019partisan}.\footnote{See
\citet{eraslan2019legislative} for a comprehensive review of multilateral
bargaining and \citet{ray2015coalition} for a review on coalition formation.}
In this paper, we extend the \citet{baron1989bargaining} model under closed
rules and add the mayor's first move and political partisanship. We consider
partisanship as a zero-sum effect that is positive for politicians aligned with
the mayor and negative otherwise. This paper investigates the often overlooked
effect of legislature size on the bargaining equilibrium strategies.

Second, this paper contributes to the literature on the effects of larger
legislatures in service provision. Following the seminal contribution by
\citet{Weingastetal1981}, the "law of $1/n$" posits that larger legislatures
increase the legislative pressure for providing targeted public goods to local
supporters at the expense of the entire polity \citep{freire2021legislature}.
This question has become more relevant since legislature size has increased
consistently over time \citep{gerring2020population}. The corollary of the "law
of $1/n$" is that increasing legislature size promotes excessive public
spending, regardless of the partisan makeup of the legislature
\citep{GilliganMatsusaka2001}.\footnote{\citet{PrimoSnyder2008} show that
legislature size may result in positive or negative expenditures, depending on
factors such as the degree that local public goods can exclude non-locals or
deadweight costs associated with taxation.} In the US metaphor, this would
explain significant infrastructure spending in smaller districts. However, the
"law of $1/n$" literature rarely mentions the relationship between public goods
provision and welfare. This paper is one of the first to study how the number
of local legislators affects welfare. Our model provides a theoretical
expectation based on the chances of electing a politician aligned with the
mayor. The empirical estimates confirm our theoretical predictions, with the
additional fact that legislature size has a null effect on expenditures in
Brazil.

Third, our article also contributes to the growing literature on municipal
partisan politics. Most papers in this field estimate the effects of the
mayor's party on political outcomes. For instance, \citet{de2016mayoral} and
\citet{gerber2011mayors} find differences in governance when comparing
Democratic and Republican mayors. Specifically about Brazil,
\citet{gouvea2021partisanship} shows that social policy expenditure increase
with left-wing mayors. \citet{frey2021legislature}, using discontinuities in
city council size after the 2009 constitutional amendment that reverted the
thresholds we study here, shows that changes in council size lower the support
for candidates backed by the municipal mayors. Our paper reinforces the view
that partisanship matters locally, even in countries with weak ideological
linkage.

Finally, this paper contributes to the literature on the causes and effects of
legislature sizes. While most papers on legislature size analyse the trade-off
between representation and efficiency \citep{frederick2008people,
jacobs2015explaining, muzzio1989size, Stigler1976}, recent articles have
explored the effects of legislature size on lawmaking and partisanship in
further detail. \citet{kirkland2014chamber} shows that larger legislatures make
it harder for politicians to know and collaborate with each other, so they rely
more heavily on party cues and committee structures. In considering the causes
of changes in legislature sizes, \citet{jacobs2015explaining} show that
population shifts increase and economic crises decrease the number of
legislators. \citet{marland2019fewer} questions the economic crises mechanism,
showing that political elites shrink legislatures mostly to signal to the
public the priority of their austerity agenda. In this article, we show that
council size affects welfare by changing the partisan composition of
the legislatures.

@@

Institutional design has a major impact on long-term welfare
\citep{acemoglu2012nations, buenodemesquita2005logic, north1989constitutions}
and democratic consolidation \citep{gandhi2019institutional,
lijphart2018institutional}. One particular problem that all polities face is
how to determine the optimal size of their legislative bodies. The literature
on this topic is extensive, and modern attempts to provide a solution to the
legislature size problem started in the early eighteenth century with
the Federalists versus Anti-Federalists debates \citep{muzzio1989size,
Stigler1976}.

The literature on legislature size highlights that the number of legislators
causes tension between representation and efficiency
\citep{frederick2008people, jacobs2015explaining, muzzio1989size,
rush2013conclusion, Stigler1976, taagepera1972size}. Larger legislatures
improve representation as both the election thresholds and the ratio of
population-per-politician decrease, which facilitate the electoral success of
ethnic and political minorities \citep{allenrepresentation,
gerring2020population}. However, better representation is often countered by
the decline in efficiency in larger legislatures \citep{Weingastetal1981}. The
enlarged size makes it harder for members to coordinate and collaborate without
partisanship and legislative structures \citep{kirkland2014chamber,
rush2013conclusion}. It also increases the potential number of veto players,
decreasing policymaking efficiency \citep{tsebelis2002veto}. Balancing these
two incentives is crucial for the optimal legislative composition
\citep{auriol2012optimal, taagepera1972size}.

This literature has often overlooked two important questions. First, does
legislature size matter in contexts where legislative bodies are not the
pivotal decision-maker? And second, what is the ultimate effect of legislature
size on welfare? These gaps are likely generated by the predominance of the
"law of 1/n" scholarship in political economy debates, which focus mainly on
the relationship between legislature size and overspending. Since
\citet{Weingastetal1981}, 30 papers have systematically tested the effects of
legislature size on government expenditures \citep{freire2021legislature}. Of
these papers, only \citet{lewis2019legislature} investigates its effects on
welfare, showing that larger legislatures have a negative welfare
effect.\footnote{\citet{koppel2004public} also shows, using a repeated game
where the proposal prerogative rotates in a legislature, that larger
legislatures increase the chances of non-cooperative outcomes, lowering public
goods provision.} \citet{frey2021legislature} analyzes the Brazilian case after
the 2009 legislature size change and finds a null effect in service provision
from 2012 to 2016. Yet, \citet{GilliganMatsusaka2001} finds that in the US
legislature size is associated primarily with higher spending on geographically
concentrated benefits, such as education and highways. These results suggest
that the legislature size and welfare nexus needs a dedicated theoretical and
empirical assessment.

Despite the importance of legislatures, the executive still dominates the
public debate on public service provision in most mayor-council (or
presidential) systems. For instance, while in 2021 the New York Times cited
Mayor Bill De Blasio 2,084 times, it mentioned the 49 seating city councilors
only 440 times in the same year. When we expand the search for the past five
years, we see that De Blasio was mentioned 8,978 while the 49 seating city
councilors were mentioned 2,046 times.\footnote{We computed these figures using
the Dow Jones Factiva Database in November 2021.} On average, the mayor was
cited four times more often than all the city councilors combined. This
executive dominance has implications for policymaking, as coalitions and
affinity with the executive are going to be crucial for service provision
\citep{buenodemesquita2005logic}.

## Model Setup

@@

The city council votes the mayor's proposal, and if councilors accept it, the
proposal is implemented. Otherwise, a reversal policy takes place. Reversal
policy in this context means that the mayor's proposal failed to secure the
majority of council votes. The decision process then leaves the mayor's office
and is transferred to the council. The types of reversal policies are crucial
for our argument, as they change the relative strength of the council when
legislators bargain with the mayor. We investigate two types of reversal
policies: A baseline reversal with no parties and a hybrid mechanism with
partisan and non-partisan concerns. The differences between these two
mechanisms are that the non-partisan councilors only care about rents, while
partisan councilors care about both rents and policy.

The mayor must convince at least half of the city councilors to support her
policy proposals. Similar to \citet{buenodemesquita2005logic}, we assume that
the selectorate, that is, the individuals who have a say in the mayor's policy
proposals, are the $N$ city councilors. The winning coalition, the minimum
number of members that must be convinced to implement the policy, is equal to
half of the selectorate ($N/2$). City councilors are motivated by policy
concerns $p$ and transfers received from the mayor or the councilor nominated
as the proposer \textbf{x}. We leave the definition of these transfers open, as
this allows us to analyze how mayors combine transfers with other incentives to
pressure city councilors. Exemples of transfers are granting electoral favors,
funding clientelistic brokers, spending in areas that the councilor has
political or personal interests, providing portfolio (municipal offices) and
public jobs within the municipal bureaucracy, or even giving campaign
contributions and bribes. The use of transfers to influence the councilors'
choices generate a bargaining cost of $C_G$. Bargaining costs vary according to
the reversal mechanism and the size of the legislature.

Finally, after learning the bargaining costs, the mayor provides a level of
public goods provision $g$ and pockets rents $r$. Public goods provision helps
the mayor to get reelected. Rents are for the mayor's direct consumption and do
not contribute to the mayor's electoral success.\footnote{This model, similar
to \citet{cowen1994rent} and \citet{buenodemesquita2005logic}, combines public
goods provision and rent extraction. For a dynamic bargaining model that
incorporates transfers to districts and public goods provision, see
\citet{battaglini2012legislative}.} This makes the mayor's expected utility a
sum of the gains from rents and the benefits from reelection. Both the
utilities from rents $u(.)$ and the probability of reelection $\pi(.)$ are
concave functions, meaning that more rents or public goods increase the utility
at a decreasing rate. The probability of reelection is multiplied by the
benefit from holding office $B_M > 0$. This benefit captures the tangible and
intangible gains that the mayor receives from holding the public office. We
assume that the benefits are high enough to rule out an equilibrium with only
rents and transfers, which is equivalent to saying that mayors strongly prefer
to be reelected. If the mayor's policy proposal is approved, then the expected
utility for the mayor is:

@@

The expected utility for the city councilors depends on the type of reversal
policy. In the baseline non-partisan reversal, when the council rejects the
mayor's proposal, a reversal stage starts with resources diminishing by a
factor of $\delta \in (0,1)$ and the random selection of one councilor. The
selected councilor becomes the next proposer. If the new proposal is accepted,
it is implemented. If councilors reject the offer again, the budget shrinks by
$\delta$ and another city councilor is nominated to propose. The process
repeats until a proposal is finally accepted.

We now consider a situation that combines both partisan and non-partisan
legislature concerns. Each councilor has a party affiliation. Party
affiliations are mutually exclusive, and if a city councilor is aligned with
the mayor's party, then she belongs to the government coalition $G \subset \{1,
2, \cdots, N \}$. Otherwise, the city councilor belongs to the opposition $O =
\{1, 2, \cdots, N \} \setminus G$. The mayor's policy generates a political
value of $p > 0$ for the councilors, and whether it adds or subtracts from the
councilor's utility depends on the party affiliation of the
councilor.\footnote{One example of partisan benefits is that, as the mayor
shares partisan ties with members of the government coalition, a fraction of
the public goods provision can be claimed by councilors as their contribution.
We propose an extension in the Online Supplementary Materials, which assumes $p
= \phi g$, where $\phi$ is the fraction of the public goods provision that can
be credit claimed by the city councilor aligned with the mayor.} If councilors
reject the mayor's proposal, no public goods are provided ($p=0$). Moreover, in
equilibrium, we assume that city councilors would prefer to receive transfers
instead of providing public goods.\footnote{In the Online Supplementary
Materials, we consider a version of the model in which public goods provision
depends on the party affiliation of the proposer so that it affects choices in
any potential history of the game. In equilibrium, the level of public goods
provision is even higher than in the model presented here, as city councilors
prefer to approve the mayor's proposal than run the risk of an opposition
member becoming the proposer.}

@@

Proposition \ref{prop1} provides our first empirically testable hypothesis:
\textit{if bargaining costs decrease when legislature size increases,
then public goods provision increases when legislature size increases}. Now
we study each reversal mechanism to determine the condition under which the
bargaining costs decrease when legislature size increases.

@@

\begin{proposition}
In the baseline non-partisan reversal, bargaining costs always increase when legislature size increases.
\end{proposition}

The intuition for the results is that every additional councilor demands more
resources from the mayor. Thus, more councilors represent higher bargaining
costs. And if the bargaining costs increase with legislature size, then any
increase in public goods provision associated with legislature size has come
from a different mechanism. Below we demonstrate that partisanship is the key
to understand how this relationship works.
